Normalization by Evaluation for Call-by-Push-Value and Polarized Lambda-Calculus

### 2019-03-26 fscd19/fscd19.tex
[ fscd19 ] arXiv version
[ fscd19 ] typos spotted by Hugo

\ccsdesc[500]{Theory of computation~Type theory}
\ccsdesc[500]{Theory of computation~Type structures}
\ccsdesc[300]{Theory of computation~Functional constructs}
\ccsdesc[300]{Theory of computation~Proof theory}
\ccsdesc[300]{Theory of computation~Categorical semantics}
\ccsdesc[100]{Theory of computation~Operational semantics}
%TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm

@@

\acknowledgements{The first author took inspiration from some
  unpublished notes by Thorsten Altenkirch titled
\emph{Another topological completeness proof for intuitionistic logic}
received by email on 16th March 2000.
Thorsten in turn credits his inspiration to an ALF proof by Thierry
Coquand.}
% \acknowledgements{ The presentation of this NbE proof and its
% Agda formalization benefited greatly from discussions with Christian
% Sattler.}%optional

%\nolinenumbers %uncomment to disable line numbering

%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

@@

In our presentation of the various lambda calculi we ignore the
concrete syntax, only consider the abstract syntax obtained by the
Curry-Howard-Isomorphism.  A term is simply a derivation tree whose
nodes are rule invocations.  Thus, a intrinsically typed,
nameless syntax is most natural,
and our syntactic classes are all presheaves over the category of
typing contexts and renamings.  The use of presheaves then smoothly
extents to the semantic constructions
\cite{catarina:csl93,altenkirchHofmannStreicher:ctcs95}.

@@

%\emph{Normalization}
Normalization is concerned with finding a normal form
$n : \Nf A \Gamma$ for each term $t : A \from \Gamma$.
The normal form should be \emph{sound},
\ie, $\tm n \cong t$
with respect to a equational theory $\cong$ on terms
(see \cref{sec:eq}).
Further, normalization should decide $\cong$, \ie,
terms $t,t'$ with $t \cong t'$ should have the same normal form $n$.
In this article, we implement only the normalization function
\fbox{$\tnorm : A \from \Gamma \to \Nf A \Gamma$}
with proving its soundness and completeness.
From a logical perspective, we will compute for each derivation of $A
\from \Gamma$ a normal derivation $\Nf A \Gamma$.

@@

All these inductive families are presheaves, due to factored
presentation using $\Add P$ and $\Cov$ the proof is not a simple
mutual induction.  Yet, in Agda, the generic proof goes through using a sized
typing for these inductive families.  Similarly, defining the
$\tjoin$ for monad $\Cov$ relies on sized typing \cite{abel:PhD}.
\[
\begin{array}{lll}
  \tjoin & : & \forall i.\
    \Cov[i]\,(\Cov[\infty]\J) \todot \Cov[\infty]\J  \\
  \tjoin[i+1]\,(\treturn[i]\;c) & = & c \\
  \tjoin[i+1]\,(\tbind^i\;t\;k)
    & = & \tbind^\infty\;
     (t : \Ne P \Gamma)\;
     \left(\stmap[\Add P]_\Gamma\;\tjoin[i]\;
       \left(k : \AddHypP P {\Cov[i]\J} \Gamma
       \right)
     \right)
\end{array}
\]
Herein, we used the sized typing of the constructors of $\Cov$:
\[
\begin{array}{lll}
  \treturn & : & \forall i.\ \J \todot \Cov[i+1] \J \\
  \tbind  & : & \forall i.\
    \tNe\,(\Comp P) \todot \Add P\,(\Cov[i]\J) \todot \Cov[i+1]\J \\
\end{array}
\]

@@

($o$) is always case tree whose leaves are neutral.

>>>>

\ccsdesc[500]{Theory of computation~Type theory}
\ccsdesc[500]{Theory of computation~Type structures}
\ccsdesc[300]{Theory of computation~Functional constructs}
\ccsdesc[300]{Theory of computation~Proof theory}
\ccsdesc[300]{Theory of computation~Categorical semantics}
\ccsdesc[100]{Theory of computation~Operational semantics}
%TODO mandatory: Please choose ACM 2012 classifications from
%https://dl.acm.org/ccs/ccs_flat.cfm

% ACM old
% F.3.2 Semantics of Programming Languages (D.3.1)
% F.4.1 Mathematical Logic (F.1.1, I.2.2, I.2.3, I.2.4)

% MSC2010
% 03F05 Cut-elimination and normal-form theorems

@@

\acknowledgements{The first author took inspiration from some
  unpublished notes by Thorsten Altenkirch titled
\emph{Another topological completeness proof for intuitionistic logic}
received by email on 16th March 2000.
Thorsten in turn credits his inspiration to an ALF proof by Thierry
Coquand.  Thanks to Hugo Herbelin for feedback and spotting some typos.}
% \acknowledgements{ The presentation of this NbE proof and its
% Agda formalization benefited greatly from discussions with Christian
% Sattler.}%optional

\nolinenumbers %uncomment to disable line numbering

\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

@@

In our presentation of the various lambda calculi we ignore the
concrete syntax, only consider the abstract syntax obtained by the
Curry-Howard-Isomorphism.  A term is simply a derivation tree whose
nodes are rule invocations.  Thus, an intrinsically typed,
nameless syntax is most natural,
and our syntactic classes are all presheaves over the category of
typing contexts and renamings.  The use of presheaves then smoothly
extents to the semantic constructions
\cite{catarina:csl93,altenkirchHofmannStreicher:ctcs95}.

@@

%\emph{Normalization}
Normalization is concerned with finding a normal form
$n : \Nf A \Gamma$ for each term $t : A \from \Gamma$.
The normal form should be \emph{sound},
\ie, $\tm n \cong t$
with respect to an equational theory $\cong$ on terms
(see \cref{sec:eq}).
Further, normalization should decide $\cong$, \ie,
terms $t,t'$ with $t \cong t'$ should have the same normal form $n$.
In this article, we implement only the normalization function
\fbox{$\tnorm : A \from \Gamma \to \Nf A \Gamma$}
with proving its soundness and completeness.
From a logical perspective, we will compute for each derivation of $A
\from \Gamma$ a normal derivation $\Nf A \Gamma$.

@@

All these inductive families are presheaves, however, due to the factored
presentation using $\Add P$ and $\Cov$, the proof is not a simple
mutual induction.  Yet, in Agda, the generic proof goes through using a sized
typing for these inductive families.  Similarly, defining the
$\tjoin$ for monad $\Cov$ relies on sized typing \cite{abel:PhD}.
\[
\begin{array}{lll}
  \tjoin & : & \forall i.\
    \Cov[i]\,(\Cov[\infty]\J) \todot \Cov[\infty]\J  \\
  \tjoin[i+1]\,(\treturn[i]\;c) & = & c \\
  \tjoin[i+1]\,(\tbind^i\;t\;k)
    & = & \tbind^\infty\;
     (t : \Ne P \Gamma)\;
     \left(\stmap[\Add P]_\Gamma\;\tjoin[i]\;
       \left(k : \AddHypP P {\Cov[i]\J} \Gamma
       \right)
     \right)
\end{array}
\]
Herein, we used the sized typing of the constructors of $\Cov$:
\[
\begin{array}{lll}
  \treturn & : & \forall i.\ \J \todot \Cov[i+1] \J \\
  \tbind  & : & \forall i.\
    \tNe\,(\Comp P) \todot \Add P\,(\Cov[i]\J) \todot \Cov[i+1]\J \\
\end{array}
\]

@@

($o$) is always a case tree whose leaves are neutral.
